### 2021-04-13 priority-gv.tex
Examples in figure.
Minor changes.
Included progress. Removed references to appendices.

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.
We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate*}[label=(\alph*)]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate*}
We will show that only \cref{ex:cycle} is typeable in PGV, while it is not typable in GV~\cite{wadler14} nor guaranteed to be deadlock-free by its predecessor~\cite{gayvasconcelos12}.

\begin{center}
\begin{minipage}{0.5\linewidth}
\begin{example}[Cyclic Structure]
  \label{ex:cycle}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new}{}
      \\
      \letpair{y}{y'}{\new}{}
      \\
      \spawn\left(%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
  }\]
\end{example}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{example}[Deadlock]
  \label{ex:deadlock}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new}{}
      \\
      \letpair{y}{y'}{\new}{}
      \\
      \spawn\left(%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
   }\]
\end{example}
\end{minipage}
\end{center}

@@

Similarly, we use syntactic sugar for the term forms of choice, which combine sending and receiving with the introduction and elimination forms for the sum and empty types. There are two constructs for binary internal choice, expressed using the meta-variable $\tm{\ell}$ which ranges over $\{\tm{\labinl},\tm{\labinr}\}$. As there is no introduction for the empty type, there is no construct for nullary internal choice:
\[
\begin{array}{l}
  \tm{\select{\ell}}\elabarrow \\
  \qquad\tm{\lambda x.\letpair{y}{z}{\new}{\andthen{\close\;(\send\;{\pair{\ell\;y}{x}})}{z}}} \\
  \tm{\offer{L}{x}{M}{y}{N}}\elabarrow \\
  \qquad\tm{\letpair{z}{w}{\recv\;{L}}{\andthen{\wait\;{w}}{\casesum{z}{x}{M}{y}{N}}}} \\
  \tm{\offeremp{L}}\elabarrow \\
  \qquad\tm{\letpair{z}{w}{\recv\;L}{\andthen{\wait\;{w}}{\absurd{z}}}}
\end{array}
\]

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
\begin{array}[t]{lcl}
  \tm{\phi}
  & \Coloneqq & \tm{\main}
    \sep        \tm{\child}
  \\
  \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  & \Coloneqq & \tm{\phi\;M}
    \sep        \tm{\ppar{\conf{C}}{\conf{D}}}
    \sep        \tm{\res{x}{x'}{\conf{C}}}
\end{array}
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \quad
  \tm{\child} + \tm{\main}  = \tm{\main}
  \quad
  \tm{\child} + \tm{\child} = \tm{\child}
  \quad
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
\]
The use of $\tm{\child}$ for child threads~\cite{lindleymorris15} overlaps with the use of the meta-variable $\cs{o}$ for priorities~\cite{dardhagay18}. Both are used to annotate sequents: flags appear on the sequent in configuration typing, and priorities in term typing. To distinguish the two symbols, they are typeset in a different font and a different colour.

@@

We must show that substitution preserves priority constraints. For this, we prove~\cref{lem:pgv-value-done}, which shows that values have finished all their communication, and that any priorities in the type of the value come from the typing environment. The proofs are by structural induction and can be found in \cref{prf:lem-pgv-value-done,prf:lem-pgv-substitution,prf:lem-pgv-subject-reduction-terms,prf:lem-pgv-subject-congruence,prf:thm-pgv-subject-reduction-confs}.
\begin{compacttheorems}
  \begin{restatablelemma}{lempgvvaluedone}%
    \label{lem:pgv-value-done}
    If $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$, then $\cs{p}=\cs{\pbot}$, and $\minpr(\ty{\Gamma})=\minpr(\ty{T})$.
  \end{restatablelemma}
  \begin{restatablelemma}{lempgvsubstitution}[Substitution]%
    \label{lem:pgv-substitution}
    \hfill\\%newline before theorem statement
    If $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$ and $\tseq[\cs{q}]{\ty{\Theta}}{V}{U'}$, then $\tseq[\cs{p}]{\ty{\Gamma},\ty{\Theta}}{\subst{M}{V}{x}}{T}$.
  \end{restatablelemma}
  \begin{restatablelemma}{lempgvsubjectreductionterms}[Subject Reduction, $\tred$]%
    \label{lem:pgv-subject-reduction-terms}
    \hfill\\%newline before theorem statement
    If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\tm{M}\tred\tm{M'}$,
    then $\tseq[\cs{p}]{\ty{\Gamma}}{M'}{T}$.
  \end{restatablelemma}
  \begin{restatablelemma}{lempgvsubjectcongruence}[Subject Congruence, $\equiv$]%
    \label{lem:pgv-subject-congruence}
    \hfill\\%newline before theorem statement
    If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\equiv\tm{\conf{C'}}$,
    then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
  \end{restatablelemma}
  \begin{restatabletheorem}{thmpgvsubjectreductionconfs}[Subject Reduction, $\cred$]%
    \label{thm:pgv-subject-reduction-confs}
    \hfill\\%newline before theorem statement
    If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\cred\tm{\conf{C'}}$,
    then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
  \end{restatabletheorem}
\end{compacttheorems}

@@

\begin{compacttheorems}
  \begin{definition}[Canonical Forms]%
    \label{def:pgv-canonical-forms}
    A~configuration $\tm{\conf{C}}$ is in canonical form if it is of the following form, where no term $\tm{M_i}$ is a value:
    \[
      \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;N)}}}.
    \]
  \end{definition}
  \begin{definition}[Normal Forms]%
    A~configuration $\tm{\conf{C}}$ is in normal form if it is of the following form, where each $\tm{M_i}$ is ready to act on $\tm{x_i}$:
    \[
      \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;V)}}}.
    \]
  \end{definition}
  \begin{restatablelemma}{lempgvcanonicalforms}[Canonical Forms]%
    \label{lem:pgv-canonical-forms}
    If $\cseq[\main]{\ty{\Gamma}}{\conf{C}}$, there exists some $\tm{\conf{D}}$ such that $\tm{\conf{C}}\equiv\tm{\conf{D}}$ and $\tm{\conf{D}}$ is in canonical form.
  \end{restatablelemma}
  \begin{restatabletheorem}{thmpgvclosedprogressconfs}[Progress, $\cred$]%
    \label{thm:pgv-closed-progress-confs}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}\equiv\conf{D}}$ for some $\tm{\conf{D}}$ in normal form.
  \end{restatabletheorem}
  \begin{proof}
    Our proof follows the reasoning by Kobayashi~\cite{kobayashi06} used in the proof of deadlock freedom for closed processes (Theorem 2). Details are in~\Cref{prf:thm-pgv-closed-progress-confs}.
  \end{proof}
  \begin{corollary}%
    \label{cor:pgv-closed-progress}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$, $\tm{\conf{C}}\centernot\cred$, and $\tm{\conf{C}}$ contains no endpoints, then $\tm{\conf{C}}\equiv\tm{\phi\;V}$ for some value $\tm{V}$.
  \end{corollary}
\end{compacttheorems}
It follows immediately from \cref{thm:pgv-closed-progress-confs} and \cref{cor:pgv-closed-progress} that a term which does not return an endpoint will complete all its communication actions, thus satisfying deadlock freedom.
\end{document}

>>>>

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.
We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate*}[label= (\alph*) ]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate*}
We will show that only \cref{ex:cycle} is typeable in PGV, while it is not typable in GV~\cite{wadler14} nor guaranteed to be deadlock-free by its predecessor~\cite{gayvasconcelos12}.

\begin{figure}
\begin{minipage}{0.5\linewidth}
\begin{example}[Cyclic Structure]%
  \label{ex:cycle}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new}{}
      \\
      \letpair{y}{y'}{\new}{}
      \\
      \spawn\left(%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
  }\]
\end{example}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{example}[Deadlock]%
  \label{ex:deadlock}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new}{}
      \\
      \letpair{y}{y'}{\new}{}
      \\
      \spawn\left(%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
   }\]
\end{example}
\end{minipage}
\end{figure}

@@

Similarly, we use syntactic sugar for the term forms of choice, which combine sending and receiving with the introduction and elimination forms for the sum and empty types. There are two constructs for binary internal choice, expressed using the meta-variable $\tm{\ell}$ which ranges over $\{\tm{\labinl},\tm{\labinr}\}$. As there is no introduction for the empty type, there is no construct for nullary internal choice:
\[
\begin{array}{lcl}
  \tm{\select{\ell}}
  & \elabarrow
  & \tm{\lambda x.\letpair{y}{z}{\new}{\andthen{\close\;(\send\;{\pair{\ell\;y}{x}})}{z}}}
  \\
  \multicolumn{3}{l}{%
  \tm{\offer{L}{x}{M}{y}{N}}\elabarrow}
  \\
  \multicolumn{3}{l}{%
  \qquad\tm{\letpair{z}{w}{\recv\;{L}}{\andthen{\wait\;{w}}{\casesum{z}{x}{M}{y}{N}}}}}
  \\
  \tm{\offeremp{L}}
  & \elabarrow
  & \tm{\letpair{z}{w}{\recv\;L}{\andthen{\wait\;{w}}{\absurd{z}}}}
\end{array}
\]

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
  \tm{\phi}
  \Coloneqq \tm{\main}
  \sep      \tm{\child}
  \hfill
  \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  \Coloneqq \tm{\phi\;M}
  \sep      \tm{\ppar{\conf{C}}{\conf{D}}}
  \sep      \tm{\res{x}{x'}{\conf{C}}}
  \hfill
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \hfill
  \tm{\child} + \tm{\main}  = \tm{\main}
  \hfill
  \tm{\child} + \tm{\child} = \tm{\child}
  \hfill
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
  \hfill
\]
The use of $\tm{\child}$ for child threads~\cite{lindleymorris15} overlaps with the use of the meta-variable $\cs{o}$ for priorities~\cite{dardhagay18}. Both are used to annotate sequents: flags appear on the sequent in configuration typing, and priorities in term typing. To distinguish the two symbols, they are typeset in a different font and a different colour.

@@

We must show that substitution preserves priority constraints. For this, we prove~\cref{lem:pgv-value-done}, which shows that values have finished all their communication, and that any priorities in the type of the value come from the typing environment.
% The proofs are by structural induction and can be found in \cref{prf:lem-pgv-value-done,prf:lem-pgv-substitution,prf:lem-pgv-subject-reduction-terms,prf:lem-pgv-subject-congruence,prf:thm-pgv-subject-reduction-confs}.
\begin{compacttheorems}
  \begin{restatablelemma}{lempgvvaluedone}%
    \label{lem:pgv-value-done}
    If $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$, then $\cs{p}=\cs{\pbot}$, and $\minpr(\ty{\Gamma})=\minpr(\ty{T})$.
  \end{restatablelemma}
  \begin{restatablelemma}{lempgvsubstitution}[Substitution]%
    \label{lem:pgv-substitution}
    \hfill\\%newline before theorem statement
    If $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$ and $\tseq[\cs{q}]{\ty{\Theta}}{V}{U'}$, then $\tseq[\cs{p}]{\ty{\Gamma},\ty{\Theta}}{\subst{M}{V}{x}}{T}$.
  \end{restatablelemma}
  \begin{restatablelemma}{lempgvsubjectreductionterms}[Subject Reduction, $\tred$]%
    \label{lem:pgv-subject-reduction-terms}
    \hfill\\%newline before theorem statement
    If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\tm{M}\tred\tm{M'}$,
    then $\tseq[\cs{p}]{\ty{\Gamma}}{M'}{T}$.
  \end{restatablelemma}
  \begin{restatablelemma}{lempgvsubjectcongruence}[Subject Congruence, $\equiv$]%
    \label{lem:pgv-subject-congruence}
    \hfill\\%newline before theorem statement
    If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\equiv\tm{\conf{C'}}$,
    then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
  \end{restatablelemma}
  \begin{restatabletheorem}{thmpgvsubjectreductionconfs}[Subject Reduction, $\cred$]%
    \label{thm:pgv-subject-reduction-confs}
    \hfill\\%newline before theorem statement
    If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\cred\tm{\conf{C'}}$,
    then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
  \end{restatabletheorem}
\end{compacttheorems}

@@

\begin{compacttheorems}
  \begin{definition}[Canonical Forms]%
    \label{def:pgv-canonical-forms}
    A~configuration $\tm{\conf{C}}$ is in canonical form if it is of the form
    \(
    \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;N)}}}
    \)
    where no term $\tm{M_i}$ is a value.
  \end{definition}
  \begin{restatablelemma}{lempgvcanonicalforms}[Canonical Forms]%
    \label{lem:pgv-canonical-forms}
    If $\cseq[\main]{\ty{\Gamma}}{\conf{C}}$, there exists some $\tm{\conf{D}}$ such that $\tm{\conf{C}}\equiv\tm{\conf{D}}$ and $\tm{\conf{D}}$ is in canonical form.
  \end{restatablelemma}
  \begin{definition}[Normal Forms]%
    A~configuration $\tm{\conf{C}}$ is in normal form if it is of the form
    \(
    \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;V)}}}
    \)
    where each $\tm{M_i}$ is ready to act on $\tm{x_i}$.
  \end{definition}
  \begin{restatabletheorem}{thmpgvclosedprogressconfs}[Progress, $\cred$]%
    \label{thm:pgv-closed-progress-confs}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}\equiv\conf{D}}$ for some $\tm{\conf{D}}$ in normal form.
  \end{restatabletheorem}
  \begin{proof}[Sketch]
    Our proof follows the reasoning by Kobayashi~\cite{kobayashi06} used in the proof of deadlock freedom for closed processes (Theorem 2). We apply \cref{lem:pgv-open-progress-terms} to each thread. Either we obtain a reduction, or each child thread is ready and the main thread ready or a value. We pick the ready term $\tm{L}$ with the smallest priority bound. If $\tm{L}$ contains new, spawn, or a link, we apply \LabTirName{E-New}, \LabTirName{E-Spawn}, or \LabTirName{E-Link}. Otherwise, $\tm{L}$ must be ready on some $\tm{x_i}$. Linearity guarantees there is some thread  $\tm{L'}$ which acts on $\tm{x'_i}$. If $\tm{L'}$ is ready, priority typing guarantees it is ready on $\tm{x'_i}$, and we apply \LabTirName{E-Send} or \LabTirName{E-Close}. If $\tm{L'}$ is not ready, it must be the main thread returning a value. We move $\tm{L}$ into the $i^{\text{th}}$ position and repeat until we either find a reduction or reach normal form.
  \end{proof}
  \begin{corollary}%
    \label{cor:pgv-closed-progress}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$, $\tm{\conf{C}}\centernot\cred$, and $\tm{\conf{C}}$ contains no endpoints, then $\tm{\conf{C}}\equiv\tm{\phi\;V}$ for some value $\tm{V}$.
  \end{corollary}
\end{compacttheorems}
It follows immediately from \cref{thm:pgv-closed-progress-confs} and \cref{cor:pgv-closed-progress} that a term which does not return an endpoint will complete all its communication actions, thus satisfying deadlock freedom.
\end{document}
